---
title: "Gentrification Analysis"
author: "Matt Lehman"
date: "`r format(Sys.time(), '%b %d, %Y')`"
output:
  html_document:
    fig_height: 4
    theme: yeti
    highlight: pygments
    toc: true
    toc_depth: 3
    toc_float: true
    df_print: kable
---

# Purpose

The purpose of this analysis is to assign gentrification values to city
neighborhoods using [Freeman's
(2003)](https://journals.sagepub.com/doi/abs/10.1177/1078087404273341)
methodsand then to verify that these assigned values are reasonable by compating
the model output to qualitative accounts.

One modification made to Freeman's defintion is that I subsituted median gross rent
for median house value. The Census estimated median house value decreased in many areas

# Packages and Setup

```{r}
knitr::opts_chunk$set(message = FALSE, dpi = 300)
```

```{r packages, echo=FALSE, warning=FALSE, message=FALSE}
library(janitor)
library(sf)
library(dplyr)
library(readr)
library(glue)
library(tibble)
library(geojsonsf)
library(ggplot2)
```

```{r}
source("geospatial_matching.R")
source("calculate_gentrification.R")
source("ggplot2_settings.R")
source("plotting_functions.R")
```

# Read and Munge Data

## Census Data

Read in 2013-2017 ACS, 2008-2012 ACS, and 2000 decennial data and munge.

### 2017

```{r}
file_path <- "C:/Users/Matt_2/Downloads/nhgis0061_csv/nhgis0061_csv"
```

Read in 2017 data and munge 


```{r 2017 msa}
msa_2017 <- read_csv(glue("{file_path}/nhgis0061_ds233_20175_2017_cbsa.csv"))
msa_2017 <- clean_names(msa_2017)

msa_2017 <- msa_2017 %>% 
  mutate(
    prop_college_degree = (ah04e021 + ah04e022 + ah04e023 + ah04e024 + ah04e025) / ah04e001
    # have to estimate 1997 - 1999 since all I have is 1990 - 1999
    ,prop_housing_built_last_20 = (ah4ze002 + ah4ze003 + ah4ze004 + (ah4ze005*(2/9))) / ah4ze001
  ) %>%
  select(gisjoin, name = cbsa, year, prop_college_degree, prop_housing_built_last_20,
         median_rent = ah5re001, median_house_value = ah53e001,
         median_household_income = ah1pe001,
         total_pop = ahy1e001) %>% 
  identity()

# filter to MSAs of interest
# make a tibble for easy reference for city abbr to MSA gisjoin

city_msa_bridge_2017 <- tibble(
  gisjoin = c("G41860", "G31080" ,"G37980", "G38300")
  ,city_abbr = c("sf", "la", "ph", "pi")
)
msa_2017 <- msa_2017 %>% 
  semi_join(city_msa_bridge_2017)
```

```{r 2017 block_grps}
block_grps_2017 <- read_csv(glue("{file_path}/nhgis0061_ds233_20175_2017_blck_grp.csv"))
block_grps_2017 <- clean_names(block_grps_2017)

block_grps_2017 <- block_grps_2017 %>% 
  mutate(
    prop_college_degree = (ah04e021 + ah04e022 + ah04e023 + ah04e024 + ah04e025) / ah04e001
    # have to estimate 1997 - 1999 since all I have is 1990 - 1999
    ,prop_housing_built_last_20 = (ah4ze002 + ah4ze003 + ah4ze004 + (ah4ze005*(2/9))) / ah4ze001
  ) %>%
  select(gisjoin, name = blkgrpa, year, prop_college_degree, prop_housing_built_last_20,
         median_rent = ah5re001, median_house_value = ah53e001,
         median_household_income = ah1pe001,
         total_pop = ahy1e001) %>% 
  identity()
```


### 2010

Read in 2008 - 2012 ACS data and munge

```{r 2012 msa}
msa_2012 <- read_csv(glue("{file_path}/nhgis0061_ds191_20125_2012_cbsa.csv"))
msa_2012 <- clean_names(msa_2012)

msa_2012 <- msa_2012 %>% 
  mutate(
    prop_college_degree = (quse021 + quse021 + quse021 + quse021 + quse021) / quse001
    # have to estimate 1992 - 1999 since all I have is 1990 - 1999
    ,prop_housing_built_last_20 = (qy1e002 + qy1e003 + (qy1e004 * (7/9))) / qy1e001
  ) %>%
  select(gisjoin, name = cbsa, year, prop_college_degree, prop_housing_built_last_20,
         median_rent = qzte001, median_house_value = qz6e001,
         median_household_income = qz6e001,
         total_pop = qspe001) %>% 
  identity()

# filter to MSAs of interest
# make a tibble for easy reference for city abbr to MSA gisjoin
city_msa_bridge_2012 <- tibble(
  gisjoin = c("G41860", "G31100" ,"G37980", "G38300")
  ,city_abbr = c("sf", "la", "ph", "pi")
)
msa_2012 <- msa_2012 %>% 
  semi_join(city_msa_bridge_2012)
```


```{r 2012 block_grps}
block_grps_2012 <- read_csv(glue("{file_path}/nhgis0061_ds191_20125_2012_blck_grp.csv"))
block_grps_2012 <- clean_names(block_grps_2012)

block_grps_2012 <- block_grps_2012 %>% 
  mutate(
    prop_college_degree = (quse021 + quse021 + quse021 + quse021 + quse021) / quse001
    # have to estimate 1992 - 1999 since all I have is 1990 - 1999
    ,prop_housing_built_last_20 = (qy1e002 + qy1e003 + (qy1e004 * (7/9))) / qy1e001
  ) %>%
  select(gisjoin, name = blkgrpa, year, prop_college_degree, prop_housing_built_last_20,
         median_rent = qzte001, median_house_value = qz6e001,
         median_household_income = qz6e001,
         total_pop = qspe001) %>% 
  identity()
```

### 2000

Read in 2000 data and munge

```{r 2000 msa}
msa_2000_a <- read_csv(glue("{file_path}/nhgis0061_ds146_2000_msa_cmsa.csv"))
msa_2000_a <- clean_names(msa_2000_a)

msa_2000_a <- msa_2000_a %>% 
  select(gisjoin, year, name = msa_cmsa, total_pop = fl5001)

msa_2000_b <- read_csv(glue("{file_path}/nhgis0061_ds151_2000_msa_cmsa.csv"))
msa_2000_b <- clean_names(msa_2000_b)

msa_2000_b <- msa_2000_b %>% 
   mutate(
    prop_college_degree = ((gkt012 + gkt013 + gkt014 + gkt015 + gkt016 +
                             gkt028 + gkt029 + gkt030 + gkt031 + gkt031) /
                             rowSums(select(., contains("gkt")), na.rm = TRUE))
    ,prop_housing_built_last_20 = (gaj001 + gaj002 + gaj003 + gaj004) / rowSums(select(., contains("gaj")), na.rm = TRUE)
  ) %>%
  select(gisjoin, prop_college_degree, prop_housing_built_last_20,
         median_rent = gbo001, median_house_value = gb7001,
         median_household_income = gmy001)

msa_2000 <- msa_2000_a %>% 
  full_join(msa_2000_b, by = "gisjoin")

# filter to MSAs of interest
# make a tibble for easy reference for city abbr to MSA gisjoin
city_msa_bridge_2000 <- tibble(
  gisjoin = c("G7362", "G4472" ,"G6162", "G6280")
  ,city_abbr = c("sf", "la", "ph", "pi")
)
msa_2000 <- msa_2000 %>% 
  semi_join(city_msa_bridge_2000)

msa_2000_a <- NULL
msa_2000_b <- NULL
```

```{r 2000 tracts}
tracts_2000_a <- read_csv(glue("{file_path}/nhgis0061_ds146_2000_tract.csv"))
tracts_2000_a <- clean_names(tracts_2000_a)

tracts_2000_a <- tracts_2000_a %>% 
  select(gisjoin, year, name = tracta, total_pop = fl5001)

tracts_2000_b <- read_csv(glue("{file_path}/nhgis0061_ds151_2000_tract.csv"))
tracts_2000_b <- clean_names(tracts_2000_b)

tracts_2000_b <- tracts_2000_b %>% 
   mutate(
    prop_college_degree = ((gkt012 + gkt013 + gkt014 + gkt015 + gkt016 +
                             gkt028 + gkt029 + gkt030 + gkt031 + gkt031) /
                             rowSums(select(., contains("gkt")), na.rm = TRUE))
    ,prop_housing_built_last_20 = (gaj001 + gaj002 + gaj003 + gaj004) / rowSums(select(., contains("gaj")), na.rm = TRUE)
  ) %>%
  select(gisjoin, prop_college_degree, prop_housing_built_last_20,
         median_rent = gbo001, median_house_value = gb7001,
         median_household_income = gmy001)

tracts_2000 <- tracts_2000_a %>% 
  full_join(tracts_2000_b, by = "gisjoin")

tracts_2000_a <- NULL
tracts_2000_b <- NULL
```

## Geospatial data

Read in neighborhoods

```{r read neigborhoods}
philly_neighborhoods_sf <- read_sf("C:/Users/Matt_2/Downloads/Neighborhoods_Philadelphia")
philly_neighborhoods_sf <- philly_neighborhoods_sf %>% 
  select(neighborhood = NAME) %>% 
  st_transform(crs = 4326)
  
# https://pittsburghpa.maps.arcgis.com/apps/OnePane/basicviewer/index.html?appid=7b284a2998454505a6f000d24ee1ded5
pitt_neighborhoods_sf <- geojson_sf("C:/Users/Matt_2/Downloads/Neighborhoods_with_SNAP_Data.geojson")
pitt_neighborhoods_sf <- pitt_neighborhoods_sf %>% 
  select(neighborhood = Neighborhood_2010_HOOD)

la_neighborhoods_sf <- geojson_sf("C:/Users/Matt_2/Downloads/Neighborhoods.geojson")
la_neighborhoods_sf <- la_neighborhoods_sf %>% 
  select(neighborhood = name)

sf_neighborhoods_sf <- geojson_sf("C:/Users/Matt_2/Downloads/Analysis Neighborhoods.geojson")
sf_neighborhoods_sf <- sf_neighborhoods_sf %>% 
  select(neighborhood = nhood)

```

Read in tracts

```{r read tracts} 
block_grp_file_path <- "C:/Users/Matt_2/Downloads/nhgis0062_shape/nhgis0062_shape"

# 2017
pa_block_grps_2017_sf <- read_sf(glue("{block_grp_file_path}/nhgis0062_shapefile_tl2017_420_blck_grp_2017"))
pa_block_grps_2017_sf <- clean_names(pa_block_grps_2017_sf) %>% 
  st_transform(crs = 4326)
ca_block_grps_2017_sf <- read_sf(glue("{block_grp_file_path}/nhgis0062_shapefile_tl2017_060_blck_grp_2017"))
ca_block_grps_2017_sf <- clean_names(ca_block_grps_2017_sf) %>% 
  st_transform(crs = 4326)


# 2012
pa_block_grps_2012_sf <- read_sf(glue("{block_grp_file_path}/nhgis0062_shapefile_tl2012_420_blck_grp_2012"))
pa_block_grps_2012_sf <- clean_names(pa_block_grps_2012_sf) %>% 
  st_transform(crs = 4326)
ca_block_grps_2012_sf <- read_sf(glue("{block_grp_file_path}/nhgis0062_shapefile_tl2012_060_blck_grp_2012"))
ca_block_grps_2012_sf <- clean_names(ca_block_grps_2012_sf) %>% 
  st_transform(crs = 4326)


# 2000
tracts_2000_sf <- read_sf("C:/Users/Matt_2/Downloads/nhgis0055_shape/nhgis0055_shape/nhgis0055_shapefile_tl2000_us_tract_2000", layer = "US_tract_2000")
tracts_2000_sf <- clean_names(tracts_2000_sf)

# PA fips is 42, CA is 06
pa_tracts_2000_sf <- tracts_2000_sf %>% 
  filter(nhgisst == "420") %>% 
  st_transform(crs = 4326)

ca_tracts_2000_sf <- tracts_2000_sf %>% 
  filter(nhgisst == "060") %>% 
  st_transform(crs = 4326)

tracts_2000_sf <- NULL
```


# Modeling Gentrification

For each city, match blocks or tracts to neighborhoods, and filter out those
that are not in neighborhoods. Weight blocks or tracts by area in neighborhood.

```{r, echo=FALSE, warning=FALSE, message=FALSE}
philly_block_grps_neighborhoods_2017_sf <- sub_geometry_to_neighborhood(philly_neighborhoods_sf, pa_block_grps_2017_sf)
philly_block_grps_neighborhoods_2012_sf <- sub_geometry_to_neighborhood(philly_neighborhoods_sf, pa_block_grps_2012_sf)
philly_tracts_neighborhoods_2000_sf <- sub_geometry_to_neighborhood(philly_neighborhoods_sf, pa_tracts_2000_sf)

pitt_block_grps_neighborhoods_2017_sf <- sub_geometry_to_neighborhood(pitt_neighborhoods_sf, pa_block_grps_2017_sf)
pitt_block_grps_neighborhoods_2012_sf <- sub_geometry_to_neighborhood(pitt_neighborhoods_sf, pa_block_grps_2012_sf)
pitt_tracts_neighborhoods_2000_sf <- sub_geometry_to_neighborhood(pitt_neighborhoods_sf, pa_tracts_2000_sf)

la_block_grps_neighborhoods_2017_sf <- sub_geometry_to_neighborhood(la_neighborhoods_sf, ca_block_grps_2017_sf)
la_block_grps_neighborhoods_2012_sf <- sub_geometry_to_neighborhood(la_neighborhoods_sf, ca_block_grps_2012_sf)
la_tracts_neighborhoods_2000_sf <- sub_geometry_to_neighborhood(la_neighborhoods_sf, ca_tracts_2000_sf)

sf_block_grps_neighborhoods_2017_sf <- sub_geometry_to_neighborhood(sf_neighborhoods_sf, ca_block_grps_2017_sf)
sf_block_grps_neighborhoods_2012_sf <- sub_geometry_to_neighborhood(sf_neighborhoods_sf, ca_block_grps_2012_sf)
sf_tracts_neighborhoods_2000_sf <- sub_geometry_to_neighborhood(sf_neighborhoods_sf, ca_tracts_2000_sf)
```

Get average measure by taking weighted averages by area and population. 


```{r, echo=FALSE, warning=FALSE, message=FALSE}
pitt_neighborhood_gentrification <- calculate_gentrification(pitt_block_grps_neighborhoods_2017_sf,
                                                             pitt_block_grps_neighborhoods_2012_sf,
                                                             pitt_tracts_neighborhoods_2000_sf,
                                                             block_grps_2017,
                                                             block_grps_2012,
                                                             tracts_2000,
                                                             msa_2017,
                                                             msa_2012,
                                                             msa_2000,
                                                             "Pittsburgh")

la_neighborhood_gentrification <- calculate_gentrification(la_block_grps_neighborhoods_2017_sf,
                                                             la_block_grps_neighborhoods_2012_sf,
                                                             la_tracts_neighborhoods_2000_sf,
                                                             block_grps_2017,
                                                             block_grps_2012,
                                                             tracts_2000,
                                                             msa_2017,
                                                             msa_2012,
                                                             msa_2000,
                                                             "Los Angeles")

sf_neighborhood_gentrification <- calculate_gentrification(sf_block_grps_neighborhoods_2017_sf,
                                                             sf_block_grps_neighborhoods_2012_sf,
                                                             sf_tracts_neighborhoods_2000_sf,
                                                             block_grps_2017,
                                                             block_grps_2012,
                                                             tracts_2000,
                                                             msa_2017,
                                                             msa_2012,
                                                             msa_2000,
                                                             "San Francisco")

philly_neighborhood_gentrification <- calculate_gentrification(philly_block_grps_neighborhoods_2017_sf,
                                                             philly_block_grps_neighborhoods_2012_sf,
                                                             philly_tracts_neighborhoods_2000_sf,
                                                             block_grps_2017,
                                                             block_grps_2012,
                                                             tracts_2000,
                                                             msa_2017,
                                                             msa_2012,
                                                             msa_2000,
                                                             "Philadelphia")
```

# Plotting Gentrification
```{r Plot Gentrification Geographic}
plot_neighborhoods_gentrification(pitt_neighborhoods_sf,
                                  pitt_neighborhood_gentrification,
                                  "Pittsburgh")

plot_neighborhoods_gentrification(philly_neighborhoods_sf,
                                  philly_neighborhood_gentrification,
                                  "Philadelphia")

plot_neighborhoods_gentrification(la_neighborhoods_sf,
                                  la_neighborhood_gentrification,
                                  "Los Angeles")

plot_neighborhoods_gentrification(sf_neighborhoods_sf,
                                  sf_neighborhood_gentrification,
                                  "San Francisco")

```

```{r}
tmp <- philly_neighborhood_gentrification %>% 
  bind_rows(pitt_neighborhood_gentrification) %>% 
  bind_rows(la_neighborhood_gentrification) %>% 
  bind_rows(sf_neighborhood_gentrification) %>% 
  group_by(city, neighborhood) %>% 
  mutate(
    gentrifiable_prior = lag(gentrifiable, order = year)
    ,gentrifying = case_when(
      !gentrifying ~ FALSE
      ,is.na(gentrifying) ~ NA
      ,gentrifiable_prior & gentrifying ~ TRUE
      ,!gentrifiable_prior & gentrifying ~ FALSE
      ,TRUE ~ NA
    )
  ) %>% 
  ungroup() %>% 
  filter(year != "2000"
         ,!is.na(gentrifying)) %>% 
  select(city, neighborhood, year, gentrifiable, gentrifying)

tmp %>% 
  group_by(city, year) %>% 
  summarise(
    neighborhoods = n()
    ,gentrifying = sum(gentrifying)
    ,pct_gentrifying =gentrifying / neighborhoods
  ) %>% 
ggplot(aes(x = year, y = pct_gentrifying, fill = city, group = city)) + 
  geom_col(position = "dodge", color = "black") +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_manual(name = "City", values = c("gray1", "gray33", "gray66", "gray99")) +
  labs(
    title = "Gentrification 2008-2012 and 2013-2017"
    ,y = "Percent of Neighborhoods Gentrifying"
    ,x = "Year"
  )
```

# Save Data for Model

```{r}
philly_neighborhood_gentrification %>% 
  bind_rows(pitt_neighborhood_gentrification) %>% 
  bind_rows(la_neighborhood_gentrification) %>% 
  bind_rows(sf_neighborhood_gentrification) %>% 
  mutate(
    year = case_when(
      year == "2000" ~ 2000
      ,year == "2008-2012" ~ 2012
      ,year == "2013-2017" ~ 2017
    ) 
  ) %>% 
  group_by(city, neighborhood) %>% 
  mutate(
    gentrifiable_prior = lag(gentrifiable, order = year)
    ,gentrifying = case_when(
      !gentrifying ~ FALSE
      ,is.na(gentrifying) ~ NA
      ,gentrifiable_prior & gentrifying ~ TRUE
      ,!gentrifiable_prior & gentrifying ~ FALSE
      ,TRUE ~ NA
    )
  ) %>% 
  ungroup() %>% 
  saveRDS("data/neighborhood_gentrification.RDS")
```


