---
title: "Gentrification Analysis"
author: "Matt Lehman"
date: "`r format(Sys.time(), '%b %d, %Y')`"
output:
  html_document:
    fig_height: 4
    theme: yeti
    highlight: pygments
    toc: true
    toc_depth: 3
    toc_float: true
---

# Purpose

The purpose of this analysis is to assign gentrification values to city
neighborhoods using [Freeman's
(2003)](https://journals.sagepub.com/doi/abs/10.1177/1078087404273341)
methodsand then to verify that these assigned values are reasonable by compating
the model output to qualitative accounts.

# Packages and Setup

```{r packages}
library(janitor)
library(sf)
library(dplyr)
library(readr)
library(glue)
library(tibble)
```

```{r}
file_path <- "C:/Users/Matt_2/Downloads/nhgis0057_csv/nhgis0057_csv"
```


# Read and Munge Data

## Census Data

Read in 2015, 2010, and 2015 data and munge.

Read in 2015 data and munge 

```{r}
file_names_2015 <- c("nhgis0057_ds215_20155_2015_cbsa", "nhgis0057_ds215_20155_2015_tract", "nhgis0057_ds216_20155_2015_cbsa", "nhgis0057_ds216_20155_2015_tract")
```

```{r 2015 msa}
msa_a_2015 <- read_csv(glue("{file_path}/{file_name}.csv", file_name = file_names_2015[1]))
msa_a_2015 <- clean_names(msa_a_2015)

msa_clean_a_2015 <- msa_a_2015 %>% 
  mutate(
    prop_college_degree = (admze021 + admze022 + admze023 + admze024 + admze025) / admze001
    # have to estimate 1995 - 2000 since all I have is 1990 - 2000
    ,prop_housing_built_last_20 = (adqse002 + adqse003 + adqse004 + (adqse005/2)) / adqse001
  ) %>%
  select(gisjoin, name = cbsa, year, prop_college_degree, prop_housing_built_last_20,
         median_house_value = adrwm001) %>% 
  identity()

msa_b_2015 <- read_csv(glue("{file_path}/{file_name}.csv", file_name = file_names_2015[3]))
msa_b_2015 <- clean_names(msa_b_2015)

msa_clean_b_2015 <- msa_b_2015 %>% 
  select(gisjoin, median_family_income = ad4lm001)

msa_2015 <- msa_clean_a_2015 %>% 
  full_join(msa_clean_b_2015, by = "gisjoin")

msa_a_2015 <- NULL
msa_clean_a_2015 <- NULL
msa_b_2015 <- NULL
msa_clean_b_2015 <- NULL

# filter to MSAs of interest
# make a tibble for easy reference for city abbr to MSA gisjoin
city_msa_bridge_2015 <- tibble(
  gisjoin = c("G41860", "G31080" ,"G37980", "G38300")
  ,city_abbr = c("sf", "la", "ph", "pi")
)
msa_2015 <- msa_2015 %>% 
  semi_join(city_msa_bridge_2015)
```

```{r 2015 tracts}
tracts_2015_a <- read_csv(glue("{file_path}/{file_name}.csv", file_name = file_names_2015[2]))
tracts_2015_a <- clean_names(tracts_2015_a)

tracts_2015_a_clean <- tracts_2015_a %>% 
  mutate(
    prop_college_degree = (admze021 + admze022 + admze023 + admze024 + admze025) / admze001
    # have to estimate 1995 - 2000 since all I have is 1990 - 2000
    ,prop_housing_built_last_20 = (adqse002 + adqse003 + adqse004 + (adqse005/2)) / adqse001
  ) %>%
  select(gisjoin, name = tracta, year, prop_college_degree, prop_housing_built_last_20,
         median_house_value = adrwm001) %>% 
  identity()

tracts_2015_b <- read_csv(glue("{file_path}/{file_name}.csv", file_name = file_names_2015[4]))
tracts_2015_b <- clean_names(tracts_2015_b)

tracts_2015_b_clean <- tracts_2015_b %>% 
  select(gisjoin, median_family_income = ad4lm001)

tracts_2015 <- tracts_2015_a_clean %>% 
  full_join(tracts_2015_b_clean, by = "gisjoin")

tracts_2015_a <- NULL
tracts_2015_a_clean <- NULL
tracts_2015_b <- NULL
tracts_2015_b_clean <- NULL
```

Read in 2010 data and munge

```{r 2010 msa}
```


```{r 2010 tracts}

```

Read in 2000 data and munge

```{r 2000 msa}
msa_2000 <- read_csv(glue("{file_path}/nhgis0056_ds151_2000_msa_cmsa.csv"))

```



## Geospatial data

Read in neighborhoods

```{r read neigborhoods}

```

Read in tracts

```{r read tracts} 

```


# Modeling Gentrification

For each city, match tracts to neighborhoods, and filter out tracts that are not in neighborhoods. Weight neighborhood means by tract population times area in neighborhood.

```{r}

```

For each city, join tract data and calculate from Freeman (2003):


1.	Have a median income less than the median for that metropolitan area at the beginning of the intercensal period.

  
2.	Have a proportion of housing built within the past 20 years lower than the proportion found at the median for the respective metropolitan area. 


3.	Have a percentage increase in educational attainment greater than the median increase in educational attainment for that metropolitan area. 


4.	Have an increase in real housing prices during the intercensal period.


```{r}
# calculate MSA median income

# compare MSA median income to mean neighborhood median in neighborhood

# calculate median proportion of housing built (MPH) within last 20 years in MSA

# calculate mean MPH for neighborhood 

# calculate median percentage increase in educational attainment for tracts in MSA

# compare median percentage increase in educational attainment

# standardize housing prices to correct for inflation and calculate changes in
# median real housing prices for neighborhoods

```

